---
title: "bringing in data.qmd"
author: "Eric W. Dewar"
format: html
editor: visual
---

## Bringing data into R

So far, we've only entered data into the console directly. There are a few other ways to bring in data from **spreadsheets** or **built-in datasets** from R itself or **generate random data**.

### **👩‍💻 Create a new script from the File menu or the new script button at the upper left. This can be a regular R script or a Quarto document.**

------------------------------------------------------------------------

## Data from your own drive using the console

### 💻 Using the desktop application

This might be the most common way that we work with data—using spreadsheets from files on our own hard drives.

#### Option 1

Let's find the **SKL-bite.csv** dataset on your drive by entering `file.choose()` into the console.

```{r}
file.choose()
```

Running this function opens a dialog box where you can specify where to find this dataset. Once you locate it, R will report the path in the console. To bring in those data, copy the file path and create an object using the `read.csv` (combine or concatenate) function. Replace the part within the parentheses.

```{r}
teeth <- read.csv(replace-with-file-path-here)

View(teeth) # previews the spreadsheet (data frame) in a tab
```

### Option 2

First, download a package called **readr** into R by entering this function into the console (or by executing this block):

```{r}
install.packages("readr")
```

Or you can use the **Install** button in the **Packages** pane and search for **readr**. You will only need to install it once.

Now go to the **Files** pane and find the dataset from your drive. Click the file name to pop up a menu—click **Import Dataset...** to bring up another dialog box. This will allow you to inspect the dataset and import it, even being able to copy the code for your script☁️ Using Posit Cloud

In the **Files** pane, select the **Upload** button. This will open a dialog box where you can select the file that you want to connect.

### ☁️ Using Posit Cloud

You can file the file on your drive using the **Upload** button in the **Files** pane. Once you have the path, you can do it like the desktop version. To bring in those data, copy the file path and create an object using the `read.csv` (combine or concatenate) function. Replace the part within the parentheses.

```{r}
teethCloud <- read.csv(replace-with-file-path-here)

View(teethCloud) # previews the spreadsheet (data frame) in a tab
```

\

------------------------------------------------------------------------

## Data from GitHub

### 💻 ☁️ For both desktop and Posit Cloud

For convenience in our class, I will frequently provide direct links to GitHub to avoid the need to fuss with data files. GitHub is a version-control system for file storage and sharing. I'll show more features later.

Navigate over to our [class repository area](https://github.com/ericdewar/biostats/tree/master) (this pop-up link opens a browser tab). Then click **SKL-bite.csv** from the file list. Clicking the file opens a page that shows a preview and other stuff. Find the **Raw** button and right-click or option click so that a menu pops up. Choose **Copy link address**.

To bring in those data, copy the URL and create an object using the `read.csv` (combine or concatenate) function, with the `url` part nested inside.

```{r}
teethGit <- read.csv(url("https://github.com/ericdewar/biostats/raw/refs/heads/master/SKL-bite.csv"))

View(teethGit) # previews the spreadsheet (data frame) in a tab
```

## 

------------------------------------------------------------------------

## Bringing in a built-in dataset

R has a number of toy datasets preinstalled. You can see the full list of them with `data()`. We'll bring in a particular one called **Orange**:

```{r}
data("Orange")
```

Once that is brought into memory, it shows up in the environment pane as a `<promise>`. Click the name name to put it with the other data objects. It will appear in the console too.\

------------------------------------------------------------------------

## Generating random samples

This function creates a vector that is a sample of 10 whole numbers taken randomly from 1 to 100 with replacement

```{r}
a <- sample(1:100, 10, replace = TRUE)
```

This makes a vector based on a normal distribution with a mean (center) of 50 and standard deviation (spread) of 5:

```{r}
n <- rnorm(n=100, mean=50, sd=5)
```

------------------------------------------------------------------------

## Using these objects in memory

The function `mean()` gives the the sample mean for an array or "column" from a data frame. We'll work through more examples later.

```{r}
mean(a) #gives the mean of the object a
```

```{r}
mean(Orange$age) #gives the sample mean for one column of a data frame
```

------------------------------------------------------------------------

## **Stuff to do**

### **Do the following and save the functions to your script:**

*Use comments* to indicate what you're doing at each step. Don't just leave uncommented code.

1.  Find the dataset called **hyena.csv** on GitHub and bring it into memory as a dataframe.

2.  View the hyena dataset as a tab.

3.  What is the mean of the ages of the hyenas in this sample?

4.  Bring in the built-in dataset called **iris**.

5.  View the iris dataset in a tab.

6.  What is the mean sepal width of all the iris sample? It's okay to ignore the species groupings for now.

7.  Generate a random sample of 50 numbers from 200 to 500 (not a normal distribution).

8.  What is the mean of this sample?
