---
title: "bringing in data.qmd"
author: "Eric W. Dewar"
format: html
editor: visual
---

## Bringing data into R {style="color: red"}

So far, we've only entered data into the console directly, but we usually don't enter data that way. Most of the time we work from spreadsheets of data that we have collected. There are a few other ways to bring in data so that we can work with them.

For class sometimes we will also work with toy datasets that are built into R itself or generate random data from probability parameters to practice working with new tools.

This tutorial is also intended to get used to working with Quarto scripts and making some basic habits of working with RStudio and its packages.

### Installing packages

R can be expanded to have new functions by installing packages. We will install many packages as we need them, but one useful set that we should install now is called the [tidyverse](https://tidyverse.org/ "tidyverse website"), a set of packages that are helpful for doing data science.

The **code block** (also called a code chunk) below has functions (commands) that will install all the tidyverse packages and then activate them. We only need to install them once. In the future we will only need to activate them when we want to use them. Pressing the green button on the upper right of the code block runs the function and opens up the bottom of the code block to show the console's output.

```{r}
install.packages("tidyverse") # installs the set of packages; comment this line out after using

library(tidyverse) # invokes/activates the packages and gives status of conflicts or superceding functions if necessary
# This function needs to be run whenever you relaunch R and want to use the package's functions
```

Once you have installed these packages, I recommend that you place a **\#** at the start of the first line so that line is ignored later.

::: callout-note
If you get an error to the effect of "could not find function," it usually means that the library that uses that function (command) isn't active or that there is a typo.

In our scripts, I will take care to activate any packages needed to run it. By default RStudio places a little banner at the top of the script if it thinks that you're trying to activate a package that isn't installed.
:::

------------------------------------------------------------------------

## Downloading scripts and files and keeping organized

### Download an example file

Most of the time we work with data we have collected ourselves using spreadsheets from files that are saved on our own hard drives.

For this tutorial, first [download this file](https://www.dropbox.com/scl/fi/icgsg4s8iskx4yn4k7v0v/SKL-bite.csv?rlkey=3fupaih592xt0kb1cuwb8vwti&dl=1) about skull length and bite force in mammal carnivores to your drive. It will probably go to your downloads folder. ⚠️ **You do not want to open this file in another application right now—just download it to your drive.**

### Setting the working directory

Next, please make a folder on your device for this class if you haven't already. You can subdivide it later if you like, but for now, at least have one folder to place all scripts and datasets.

::: callout-important
I **strongly** recommend that you keep files related to projects organized. Please do not run this whole class from your downloads folder or your desktop.
:::

Once you download scripts or datasets to your drive, it would make sense to move them to a folder where you will keep everything related to this class (or to a particular project). You can avoid some work later if you then **set the working directory** to tell R where to look for your files first. There are three ways to do this.

1.  From the top menu bar go to **Session** \> **Set Working Directory** \> **Choose Directory...** and then choose the location from the dialog box

2.  From the **Files** pane check the box next to the folder you want and then click the gear ⚙️ on the upper right of the pane. Choose **Set As Working Directory** from the drop-down menu.

3.  For [later]{.underline}: the most repeatable way is to use the `setwd` function the console or a code block with the file path (more on finding the file path below).

### Bringing in scripts

You can bring in scripts like this (1) from the **Files** pane by clicking on it, (2) from the **File** menu above, or (3) from the little folder 📂 icon in the ribbon above or the little arrow next to it.\
\

------------------------------------------------------------------------

## Loading in data from your own drive using the console

::: callout-important
Choose [**either**]{.underline} the desktop RStudio IDE way or the Post Cloud way below, depending on how you are working.
:::

### 💻 Bringing in data with the desktop IDE

#### 1. Using `file.choose` in the console

Let's find the **SKL-bite.csv** dataset on your drive by entering `file.choose()` into the console. Running this function opens a dialog box where you can specify where to find this dataset. Once you locate it, R will report the path in the console.

```{r}
file.choose() # find the file you want
```

To bring in those data, copy the file path [including the quotation marks]{.underline} and create an object called `teeth` using the `read.csv` (combine or concatenate) function. [Replace]{.underline} the part within the parentheses.

```{r}
teeth <- read.csv(insert-file-path-in-quotes-here) # create the object in working memory
```

You will see the object teeth in the **Environment** pane now. If you click the little blue arrow to the left of the name it will give some details about the variables in the data frame (the R word for a spreadsheet-like table). There are a few ways to inspect the data: the `head` function will show the first few rows and the `View` function (note the [capital V]{.underline}) opens up the data frame in another tab. This new tab will open up over this script window. The **summary** button there will open a few graphics to describe your variables too. [Here](https://sydney-informatics-hub.github.io/lessonbmc/02-BMC_R_Day1_B/index.html) is additional info about variable classes if you're curious.

```{r}
head(teeth) # checks the first few rows

View(teeth) # opens data in a new tab
```

#### 2. Using the `readr` package through the Files pane

We can also use the package `readr` from the tidyverse suite.

Go to the **Files** pane and find the dataset from your drive. Click the file name to pop up a menu and select **Import Dataset...** to bring up another dialog box. This will allow you to inspect the dataset and import it, even being able to copy the code for your script.

### ☁️ Bringing in data with Posit Cloud

In Posit Cloud, you can bring the file on your drive into R's active memory using the **Upload** button in the **Files** pane. Once you have the path, you can do it like the desktop version. To bring in those data, copy the file path and create an object using the `read.csv` (combine or concatenate) function. Replace the part below within the parentheses.

```{r}
teeth <- read.csv(insert-file-path-in-quotes-here) # create the object in working memory
```

You will see the object teeth in the **Environment** pane now. If you click the little blue arrow to the left of the name it will give some details about the variables in the data frame (the R word for a spreadsheet-like table). There are a few ways to inspect the data: the `head` function will show the first few rows and the `View` function (note the [capital V]{.underline}) opens up the data frame in another tab. This new tab will open up over this script window. The **summary** button there will open a few graphics to describe your variables too. [Here](https://sydney-informatics-hub.github.io/lessonbmc/02-BMC_R_Day1_B/index.html) is additional info about variable classes if you're curious.

```{r}
head(teeth) # checks the first few rows

View(teeth) # opens data in a new tab
```

\

------------------------------------------------------------------------

## Data from GitHub

### 💻 ☁️ For both desktop and Posit Cloud

For convenience in our class, I will frequently provide direct links to data housed on GitHub to avoid the need to fuss with data files. GitHub is a version-control system for file storage and sharing. I'll show more features later.

Navigate over to our [class repository area](https://github.com/ericdewar/biostats/tree/master). Then click **birds.csv** from the file list. Clicking the file opens a page that shows a preview and other stuff. Find the **Raw** button and right-click or option click so that a menu pops up. Choose **Copy link address**.

To bring in those data, you would copy the URL and create an object using the `read.csv` (combine or concatenate) function, with the `url` part nested inside and quotes around the link, as shown in this block. Usually I will just provide the link that I want you to use in the scripts that I share.

```{r}
birbs <- read.csv(url("https://github.com/ericdewar/biostats/raw/refs/heads/master/birds.csv"))

View(birbs) # previews the spreadsheet (data frame) in a tab
```

------------------------------------------------------------------------

## Bringing in a built-in dataset

R has a number of toy datasets preinstalled. You can see the full list of them with `data()`. We'll bring in a particular one called **Orange**:

```{r}
data("Orange")
```

Once that is brought into memory, it shows up in the environment pane as a `<promise>`. Click the name to put it with the other data objects. It will be reported in the console too.\
\

------------------------------------------------------------------------

## Clearing datasets from memory

I try to clear things out when I'm not working with them. This includes tabs and datasets and recent functions in the console. I don't worry about losing work because we use scripts to help us to keep our place and recreate what we need when we are continuing our work doing similar work in the future.

To clear the files out of the working memory click the little broom icon 🧹 in the **Environment** pane. **Go ahead and do that now, and then click to confirm when prompted.** Sweeping out the objects there does not delete any files from your drive; it just removes them from working memory.

You can also clear out the console the same way. Everything that you have put in has been saved in the **History** pane even if you don't have it in a script.\
\

------------------------------------------------------------------------

## Generating random samples

We can also create datasets of random numbers. This function creates an data object called a vector that is a sample of 10 whole numbers taken randomly from 1 to 100 with replacement.

```{r}
a <- sample(1:100, 10, replace = TRUE)  

a 
# just entering the object name reports it in the console
```

This makes a vector based on a normal distribution with a mean (center) of 50 and standard deviation (spread) of 5:

```{r}
n <- rnorm(n=100, mean=50, sd=5)

n
```

------------------------------------------------------------------------

## Using these objects in memory

The function `mean()` gives the the sample mean for a series of numbers that R calls an array or "column" from a data frame. We'll work through more examples later.

```{r}
mean(a) # gives the mean of the object a
```

```{r}
data("Orange") # bring the dataset back in, assuming that the active objects were cleared from the Environment

mean(Orange$age) # gives the sample mean for one column of a data frame
```

------------------------------------------------------------------------

## Workshop practice

If there is time in class, download the file called **2026-09-15 workshop.qmd** from the current Canvas module. Then open it and work through the activities. We will decide in class if this will be submitted this morning or tonight.
